---
title: "Data Preparation"
author: "Nate Breznau"
date: "6/4/2021"
output: html_document
---

```{r setup, include=FALSE}

pacman::p_load("tidyverse", "countrycode", "zoo", "ggplot2")
```


### Load Data

Data files were worked up from raw data in the folder "data/data_prep_routines". There are many more data files here than used in the blog post in case users are interested.

```{r files, warning = F, message = F}

deaths <- read_csv(here::here("data","deaths_long.csv"))
geip <- read_csv(here::here("data","geip_ilo.csv"))
gov_int <- read_csv(here::here("data","gov_int_blavatnik.csv"))
hours_lost <- read_csv(here::here("data","hours_lost_ilo.csv"))
polity2 <- read_csv(here::here("data","polity2.csv"))
sentiment <- read_csv(here::here("data","sentiment_rp.csv"))
socx_owd <- read_csv(here::here("data","socx_owd.csv"))
wdi <- read_csv(here::here("data","wdi.csv"))
fiw <- read_csv(here::here("data","fiw.csv"))
cis <- read_csv(here::here("data","cis.csv"))
gbp <- read_csv(here::here("data","gbp.csv"))
gini <- read_csv(here::here("data","gini_solt.csv"))

```

### Merge Data

```{r merge, warning = F, message = F}
covid <- full_join(deaths, gov_int, by = c("iso3c","day"))
covid <- full_join(covid, sentiment, by = c("iso3c","day"))
covid <- covid %>%
  left_join(cis, by = c("iso3c","day")) %>%
  left_join(gbp, by = c("iso3c","day")) %>%
  left_join(geip, by = c("iso3c")) %>%
  left_join(hours_lost, by = c("iso3c")) %>%
  left_join(polity2, by = c("iso3c")) %>%
  left_join(wdi, by = c("iso3c")) %>%
  left_join(socx_owd, by = c("iso3c")) %>%
  left_join(fiw, by = c("iso3c")) %>%
  left_join(gini, by = c("iso3c"))


# country name column
covid <- covid %>%
  mutate(country = countrycode(iso3c, "iso3c", "country.name")) %>%
  select(iso3c, country, everything())
```

### Interpolate

Where possible

Additions:
(Guesstimation)
1. Polity2 
ABW, PRI, BMU, FRO, GUM & VIR = 7 (independent democracies, technical ownership from other country); VAT = 10
2. Employment Protection Coverage (lfcov)
QAT, ARE = 55; IRQ, AFG, BIH, MKD, SRB, MNE = 45; PSE, MNG, TJK, MOZ, KHM = 20; SOM, ERI, TLS, SSD, TON = 3; MDV, COM, GUM, PRI, VUT, VIR, ABW, BMU, MHL, ATG, GRL = 65
('Google'/Wikipedia)
2. GDP 
SOM = 150; SYR = 2000
3. Unemp
AND, LIE = 4; DMA, FRO, KNA, ATG = 7
4. Fertility
DMA, KNA = 1.9; MCO = 1.5

```{r interp, message = F, warning = F}

# replace missing polity2 with fiw
# update GDP
# combine national and ILO unemployment scores, then update
covid <- covid %>%
  mutate(polity2_i = ifelse(!is.na(polity2), polity2, fiw),
         polity2_i = ifelse(iso3c %in% c("ABW", "PRI", "BMU", "FRO", "GUM", "VIR"), 7, 
                            ifelse(iso3c == "VAT", 10, polity2_i)),
         gdp_cap = ifelse(iso3c == "SOM", 150,
                   ifelse(iso3c == "SYR", 2000, gdp_cap)),
         unemp_i = ifelse(!is.na(unemp_nat), unemp_nat, unemp_ilo),
         unemp_i = ifelse(iso3c %in% c("AND", "LIE"), 4,
                   ifelse(iso3c %in% c("DMA", "FRO", "KNA", "ATG"), 7, unemp_i)),
         fertility_i = ifelse(!is.na(fertility), fertility, 
                       ifelse(iso3c %in% c("DMA","KNA"), 1.9, 
                       ifelse(iso3c == "MCO", 1.5, NA))),
         lfcov_i = ifelse(!is.na(lfcov), lfcov, 
                          ifelse(iso3c %in% c("QAT", "ARE"), 55,
                          ifelse(iso3c %in% c("IRQ", "AFG", "BIH", "MKD", "SRB", "MNE"), 45,
                          ifelse(iso3c %in% c("PSE", "MNG", "TJK", "MOZ", "KHM", "FSM", "GNB", "SUR"), 20,
                          ifelse(iso3c %in% c("SOM", "ERI", "TLS", "SSD", "TON"), 3,
                          ifelse(iso3c %in% c("MDV", "COM", "GUM", "PRI", "VUT", "VIR", 
                                              "ABW", "BMU", "MHL", "ATG", "GRL", "AND", "FRO"), 65, NA)))))),
         deaths = ifelse(is.na(deaths) & day < as.Date("2020-02-28"), 0, deaths),
         cases = ifelse(is.na(cases) & day < as.Date("2020-02-28"), 0, cases))

# find missing country-data
missing <- covid %>%
  group_by(iso3c) %>%
  summarise_all(mean) %>%
  mutate(country = countrycode(iso3c, "iso3c", "country.name")) %>%
  select(-c(day:H8_elderly_protection))

# fill in socx with prediction from GDP cap and lfcov
m_socx <- lm(socx_owd ~ log(gdp_cap) + lfcov_i, data = missing)
missing$socx_p <- predict(m_socx, newdata = missing)

missing <- missing %>%
  mutate(socx_owd = ifelse(!is.na(socx_owd), socx_owd, socx_p),
         gov_prot = scale(socx_owd*lfcov_i)) %>% # make social protection scale 
  select(iso3c, gov_prot, socx_p)

covid <- covid %>%
  left_join(missing, by = c("iso3c"))


```


### Infection Rate + Data Transformations

Calculated as deaths per capita with an 18-day lag, adjusted for age structure from:

O’Driscoll, Megan et al. 2021. “Age-Specific Mortality and Immunity Patterns of SARS-CoV-2.” Nature 590(7844): 140–45.

The average below 65 is roughly 0.05% and the average for above 65 is roughly 2.5%. They find that under 65, the cross-country comparability in IFR is very reliable. This means over 65 deaths are 50:1. This means that the deaths need to be adjusted for 50/1 * ratio of over 65 to not.

```{r ifr, message = F, warning = F}
# the behaviors and concerns data do not perfectly match, this designs a value to capture the behaviros in the past 7 days prior to the first concerns data observation to allow for a scatterplot of many more countries
covid_x <- covid %>%
  subset(day < as.Date("2020-03-31") & day > as.Date("2020-03-23")) %>%
  group_by(iso3c) %>%
  arrange(day) %>%
  summarise(behave_fill = mean(behave, na.rm = T)) %>%
  mutate(day = as.Date("2020-03-31")) %>%
  subset(behave_fill!= "NaN")

covid <- covid %>%
  left_join(covid_x, by = c("iso3c","day"))

covid <- covid %>%
  group_by(iso3c) %>%
  arrange(day) %>%
  mutate(deaths_day = deaths - lag(deaths),
         deaths_day_6roll = zoo::rollmean(deaths_day, k = 6, fill = NA), # 6 day rolling average means that every estimate includes one Sunday (the day reporting causes variation)
         deaths_6roll = zoo::rollmean(deaths, k = 6, fill = NA),
         death_lag18 = lead(deaths_6roll, n = 18), 
         death_day_lag18 = lead(deaths_day_6roll, n = 18),
         deaths_young = deaths*((1/50)*((100-pop65)/100)),
         deaths_day_young = deaths_day_6roll*((1/50)*((100-pop65)/100)),
         deaths_old = deaths*((49/50)*(pop65/100)),
         deaths_day_old = deaths_day_6roll*((49/50)*(pop65/100)),
         infection_young = deaths_young*(100/.05), # young infection rate = IFR * young deaths
         infection_day_young = deaths_day_young*(100/.05),
         infection_old = deaths_old*(100/2.5), # old infection rate = IFR * old deaths
         infection_day_old = deaths_day_old*(100/2.5),
         infection_total = infection_young + infection_old,
         infection_day = infection_day_young + infection_day_old,
         infection_pc = (infection_total/pop)*10000, # infections per day 10,000
         infection_pc_3roll = zoo::rollmean(infection_pc, k = 5, fill = NA),
         infection_pc18 = lead(infection_pc_3roll, n = 18),
         infection_day_pc = (infection_day/pop)*10000, # infections per day 10,000
         infection_day_pc_3roll = zoo::rollmean(infection_day_pc, k = 5, fill = NA),
         infection_day_pc18 = lead(infection_day_pc_3roll, n = 18),
         deaths_pc = (deaths/pop)*10000, # deaths per 10,000
         region = countrycode(iso3c, "iso3c","un.region.name"),
         subregion = countrycode(iso3c, "iso3c", "un.regionsub.name"),
         infection_day_delta = infection_day_pc18 - lag(infection_day_pc18),
         infection_week_delta = (infection_day_pc18 + lag(infection_day_pc18) + lag(infection_day_pc18, 2) + lag(infection_day_pc18, 3))/4 - (lag(infection_day_pc18,6) + lag(infection_day_pc18,7) + lag(infection_day_pc18, 8) + lag(infection_day_pc18, 9))/4, # 'weekly' change in infection rate (actually 4 days)
         cases_pc = (cases/pop)*10000, # confirmed cases per 10000
         cases_week = zoo::rollmean(cases_pc - lag(cases_pc, 7), k = 4, fill = NA),
         cases_week_delta = zoo::rollmean((cases_week - lag(cases_week, 5)), k = 4, fill = NA),
         behave_3roll = zoo::rollmean(behave, k = 3, fill = NA), # to help fill in some missing days and smooth
         behave_3roll_prefill = ifelse(day == as.Date("2020-03-31") & is.na(behave_3roll), behave_fill, behave_3roll),
         iso_period = ifelse(!is.na(period), paste0(iso3c,period), NA)) %>% #set up period counter
  select(iso3c, day, deaths, deaths_day:cases_week_delta, pop, everything(), -behave_fill)
```



```{r out}
saveRDS(covid, here::here("data","covid.RDS"))
```
